Locality and Effect Reflection

### 2025-07-09 paper.tex
Some edits

Note that the locality of the handler value ensures that these
operations can only be performed when they are available in the ambient
monad. This provides the same guarantees as using an effect system.

@@

The semantics is defined in terms of presheaves of the opposite of the
category of \emph{bounded Lawvere theories}. Bounded Lawvere theories
are a categorical representation of algebraic theories: we represent the
ambient algebraic monad by the corresponding algebraic theory. The key
constructs of effect reflection are then defined as follows:

@@

%\begin{lstlisting}[style=oxcaml]
%module Half_reflection (L : Op) (R : Op) : sig
%  module L_eff = Reflection(L)
%  module R_eff = Reflection(R)
%  module Sum_eff = Reflection(Sum(L)(R))
%  module Proj = Project(L)(R)
%
%  let reify (f : Handler(L).t @ local -> Handler(R).t @ local -> 'a) lh =
%    let rec loop lh = function
%      | Term(Sum(L)(R)).Return ->
%          Half_term(L)(R).Return v
%      | Term(Sum(L)(R)).Op(Left op, k) ->
%          loop lh (k (L_eff.perform op lh))
%      | Term(Sum(L)(R)).Op(Right op, k) ->
%          Half_term(L)(R).Op(op, (fun res lh -> loop lh (k res)))
%    in
%    loop (Sum_eff.reify (fun h -> f (Proj.outl h) (Proj.outr h))) lh
%
%  let rec reflect (f : Handler(L).t @ local -> 'a Half_term(L)(R).t) lh rh =
%    match f lh with
%    | Return x -> x
%    | Op(op, k) -> reflect (k (R_eff.perform op rh)) lh rh
%end
%\end{lstlisting}

>>>>

Note that the locality of the handler value ensures that these
operations can only be performed when they are available in the ambient
monad. This provides the same guarantees as using an effect system
without any type system machinery beyond tracking locality.

@@

The semantics is defined in terms of presheaves of \emph{bounded Lawvere
  theories}. Bounded Lawvere theories are a categorical representation
of algebraic theories: we represent the ambient algebraic monad by the
corresponding algebraic theory. The key constructs of effect reflection
are then defined as follows:

@@

%\begin{lstlisting}[style=oxcaml]
%  module Half_reflection (L : Op) (R : Op) : sig
%    module L_eff = Reflection(L)
%    module R_eff = Reflection(R)
%    module Sum_eff = Reflection(Sum(L)(R))
%    module Proj = Project(L)(R)
%
%    let reify (f : Handler(L).t @ local -> Handler(R).t @ local -> 'a) lh =
%      let rec loop lh = function
%        | Term(Sum(L)(R)).Return ->
%            Half_term(L)(R).Return v
%        | Term(Sum(L)(R)).Op(Left op, k) ->
%            loop lh (k (L_eff.perform op lh))
%        | Term(Sum(L)(R)).Op(Right op, k) ->
%            Half_term(L)(R).Op(op, (fun res lh -> loop lh (k res)))
%      in
%      loop (Sum_eff.reify (fun h -> f (Proj.outl h) (Proj.outr h))) lh
%
%    let rec reflect (f : Handler(L).t @ local -> 'a Half_term(L)(R).t) lh rh =
%      match f lh with
%      | Return x -> x
%      | Op(op, k) -> reflect (k (R_eff.perform op rh)) lh rh
%end
%\end{lstlisting}
